---
layout: topic
title: Before we start (notes)
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

## R and RStudio

### R

- Free and open source general statistical programming language
- Broad statistical and graphics functions
- Huge community of add-on packages
- Not just for statistics, but also general purpose programming.

### RStudio

- Integrated Development Environment (IDE) for R
- Separate program, developed by a company but also free and open
  source.
- By far the best thing to use when learning R.

### RStudio orientation

Four panes (can be re-organized, but always these four panes)

- Scripts (top-left)
- R console (bottom-left)
- Environment/history (top-right)
- Files/plots/help/etc (bottom-right)

## R as a calculator

- `Ctrl-Shift-2` to focus only on the R console.
- `Ctrl-Shift-0` to get back.

- Explain `>` prompt and do some calculations

```{r add, results="markup"}
2016 - 1969
```

```{r calculations, eval=FALSE}
4*6
4/6
4^6
log(4)
log10(4)
log(4, 10)
log(4, 6)
```

- Explain that `log` is a function with two arguments.

## Getting help

- Pause for a moment after typing `log`
- Type `?log`

## Need for scripts

- Want to save what we're doing
- Really want to organize the scripts and data for a project

### Create a new R project

- File → New Project → New directory → Empty project
- Give it a name, like `~/data-carpentry`
- Under files tab on the bottom right, click "New Folder" and create a
  folder named `data`

### Create an R script

- File → New File → R script
- Save it in the current working directory, `data-carpentry-script.R`

## Typing into the script and pasting to console

- Type some stuff into the script

```{r rscript}
# R intro

2016 - 1969
```

- Paste to console with <kbd>`Ctrl`</kbd> + <kbd>`Enter`</kbd>

- Explain comment sign, `#`

<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
