---
layout: topic
title: Intro to R
author: Data Carpentry contributors
minutes: 45
---

```{r knitr_options, echo=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

------------

> ## Learning Objectives
>
> * load external data (CSV files) in memory using the survey table
>  (`surveys.csv`) as an example
> * explore the structure and the content of a data frame in R
> * understand what factors are and how to manipulate them
> * understand the concept of a `data.frame`
> * use sequences
> * know how to access any element of a `data.frame`

------------

## Basics of R

R is a versatile, open source programming/scripting language that's useful both
for statistics but also data science. Inspired by the programming language S.

* Free/Libre/Open Source Software under the GPL.
* Superior (if not just comparable) to commercial alternatives. R has over 7,000
  user contributed packages at this time. It's widely used both in academia and
  industry.
* Available on all platforms.
* Not just for statistics, but also general purpose programming.
* For people who have experience in programmming: R is both an object-oriented
  and a so-called [functional language](http://adv-r.had.co.nz/Functional-programming.html).
* Large and growing community of peers.


## Presentation of RStudio

Let's start by learning about RStudio, the Integrated Development Environment
(IDE) that we will use to write code, navigate the files found on our computer,
inspect the variables we are going to create, and visualize the plots we will
generate. RStudio can also be used for other things (e.g., version control,
developing packages) that we will not cover during the workshop.

RStudio is divided into 4 "Panes": the editor for your scripts and documents
(top-left), the R console (bottom-left), your environment/history (top-right),
and your files/plots/packages/help/viewer (bottom-right). The placement of these
panes and their content can be customized.


## R as a calculator

The lower-left pane (the R "console") is where you can interact with R
directly. The `>` sign is the R "prompt". It indicates that R is
waiting for you to type something.

You can type <kbd>`Ctrl`</kbd> + <kbd>`Shift`</kbd> + <kbd>`2`</kbd>
to focus just on the R console pane. Use <kbd>`Ctrl`</kbd> +
<kbd>`Shift`</kbd> + <kbd>`0`</kbd> to get back to the four panes. I
use this when teaching but not otherwise.

Let's start by subtracting a couple of numbers.

```{r add, results='show'}
2016 - 1969
```

R does the calculation and prints the result, and then you get the `>`
prompt again. We won't discuss what that number might mean. (The `[1]`
in the results is a bit weird; you can ignore that for now.)

You can use R as a calculator in this way.

```{r calculations, eval=FALSE}
4*6
4/6
4^6
log(4)
log10(4)
```

Here `log` is a _function_ that gives you the natural logarithm. Much of
your calculations in R will be through functions. The value in the
parentheses is called the function "argument". `log10` is another
function, that calculates the log base 10. There can be more than
one argument, and some of them may have default values. `log` has a
second argument `base` whose default (the value if left unspecified)
is _e_.

### Getting help

If you type `log` and pause for a moment, you'll get a pop-up with
information about the function.

Alternatively, you could type

```{r help, eval=FALSE}
?log
```

and the documentation for the function will show up in the lower-right
pane. These are often a bit _too_ detailed, and so they take some
practice to read. I generally focus on Usage and Arguments, and then
on Examples at the bottom. We'll talk more about getting help later.

## Need for Scripts

We can go along, typing directly into the R console. But there won't
be an easy way to keep track of what we've done.

It's best to write R "scripts" (files with R code), and work from them.
And when we start creating scripts, we need to worry about how we
organize the scripts and data for a project.

And so let's pause for a moment and talk about file organization.


## Creating an R Project

It is good practice to keep a set of related data, analyses, and text
self-contained in a single folder, called the **working directory**. All of the
scripts within this folder can then use *relative paths* to files that indicate
where inside the project a file is located (as opposed to absolute paths, which
point to where a file is on a specific computer). Working this way makes it
a lot easier to move your project around on your computer and share it with
others without worrying about whether or not the underlying scripts will still
work.

RStudio provides a helpful set of tools to do this through its "Projects"
interface, which can create a working directory for you (or use an
existing one) and also
remembers its location (allowing you to quickly navigate to it) and optionally
preserves custom settings and open files to make it easier to resume work after
a break. Below, we will go through the steps for creating an RProject for this
tutorial.

We assume you've already got a `DataCarpentry` folder that you're
going to use.

* Start RStudio (presentation of RStudio -below- should happen here)
* Under the `File` menu, click on `New project`, choose `Existing directory`, then
  click the "`Browse`" button and find your `DataCarpentry` folder.
* Click on "Create project"
* Create a new R script (File > New File > R script) and save it in
  your `Code` subfolder (e.g. `~/DataCarpentry/Code`)




## Interacting with R

While you can type R commands directly at the `>` prompt in the R
console, I recommend typing your commands into a script, which you'll
save for later reference, and then executing the commands from there.

Start by typing the following into the R script in the top-left pane.

```{r rscript}
# R intro

2016 - 1969
```

Save the file clicking the computer disk icon, or by typing
<kbd>`Ctrl`</kbd> + <kbd>`S`</kbd>.

Now place the cursor on the line with `2016 - 1969` and type
<kbd>`Ctrl`</kbd> + <kbd>`Enter`</kbd>. The command will be copied to
the R console and executed, and then the cursor will move to the next
line.

You can also highlight a bunch of code and execute the block all at
once with <kbd>`Ctrl`</kbd> + <kbd>`Enter`</kbd>.


### Commenting

Use `#` signs to comment. Anything to the right of a `#` is ignored by R,
meaning it won't be executed. Comments are a great way to describe what your
code does within the code itself, so comment liberally in your R scripts.

## Assignment operator

We can assign names to numbers and other objects with the assignment
operator, `<-`. For example:

```{r assign}
age <- 2016-1969
```

Type that into your script, and use <kbd>`Ctrl`</kbd> +
<kbd>`Enter`</kbd> to paste it to the console.

You can also use `=` as assignment, but that symbol can have other
meanings, and so I recommend sticking with the `<-` combination.

In RStudio, typing <kbd>Alt</kbd> + <kbd>-</kbd> will write ` <- ` in
a single keystroke.

If you've assigned a number to an object, you can then use it in
further calculations:

```{r sqrt, results='show'}
sqrt(age)
```

You can assigned _that_ to something, and then use it.

```{r round_sqrt_age}
sqrt_age <- sqrt(age)
round(sqrt_age)
round(sqrt_age, 2)
```
## Object names

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too
long.

They cannot start with a number (`2x` is not valid, but `x2` is).

R is case sensitive
(e.g., `weight_kg` is different from `Weight_kg`). There are some names that
cannot be used because they are the names of fundamental functions in R (e.g.,
`if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list).

In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). In doubt
check the help to see if the name is already in use.

It's also best to avoid
dots (`.`) within a variable name as in `my.dataset`. There are many functions
in R with dots in their names for historical reasons, but because dots have a
special meaning in R (for methods) and other programming languages, it's best to
avoid them. It is also recommended to use nouns for variable names, and verbs
for function names. It's important to be consistent in the styling of your code
(where you put spaces, how you name variable, etc.). In R, two popular style
guides are [Hadley Wickham's](http://adv-r.had.co.nz/Style.html) and
[Google's](https://google.github.io/styleguide/Rguide.xml).

### Using named arguments

For the `round()` function the 1st argument is the number to round and the 2nd is how many digits to round to. There are some functions that take many arguments and you can imagine that it might get confusing trying to keep them in order. In that case, it is often better to expiclity name the arguments. You can find the argument names using the help methods we discussed earlier.

Here is an example using named arguments with round.  When the arguments are named the order doesn't matter. You might enter the first few important arguments positionally, and later ones by naming them.

```{r round_w_named_args}
round(x = sqrt_age, digits = 2)
round(digits = 2, x = sqrt_age)
round(sqrt_age, digits = 2)
```

In the coming materials we may enter named arguments, as needed.

### Challenge

What is the value of `y` after doing the following?

```{r, eval=FALSE}
x <- 50
y <- x * 2
x <- 80
```

<!-- end challenge -->

Objects don't get linked to each other, so if you change one, it won't
affect the values of any others.

## Plunge straight into the Survey Data

We will continue to look at the species and weight of animals caught
in plots in a study area in Arizona over time. The dataset is stored
as a CSV file: each row holds information for a single animal, and the
columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

The data are available at <http://kbroman.org/datacarp/portal_data_joined.csv>.

We can read that data straight from the web, as follows

```{r read_csv_from_web}
surveys <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```


### Objects in your workspace

The objects you create get added to your "workspace". You can list the
current objects with `ls()`.

```{r ls}
ls()
```

RStudio also shows the objects in the Environment panel.

## Data frames

The data are stored in what's called a "data frame". It's a big
rectangle, with rows being observations and columns being
variables. The different columns can be different types (numeric,
character, etc.), but they're all the same length.

Use `head()` to view the first few rows.

```{r head}
head(surveys)
```

Use `tail()` to view the last few rows.

```{r tail}
tail(surveys)
```

Use `str()` to look at the structure of the data.

```{r str}
str(surveys)
```

This shows that there are `r nrow(surveys)` rows and `r ncol(surveys)`
columns, and then for each column, it gives information about the data
type and shows the first few values.

For these data, the columns have two types: integer, or "Factor".
Factor columns are text with a discrete set of possible values. We'll
come back to this in a bit.

### Challenge

Study the output of `str(surveys)`. How are the missing values being treated?

<!-- end challenge -->

### Another summary

Another useful function in `summary()`.

```{r summary}
summary(surveys)
```

For the numeric data, you get six statistics (min, 25th percentile,
median, mean, 75th percentile, and max). For the factors, you get a
table with counts for the most-frequent "levels", and then for the rest.


## Inspecting data frames

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a `data.frame`. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data.

* Size:
    * `dim()` - returns a vector with the number of rows in the first element,
      and the number of columns as the second element (the __dim__ensions of the
      object)
    * `nrow()` - returns the number of rows
    * `ncol()` - returns the number of columns

* Content:
    * `head()` - shows the first 6 rows
    * `tail()` - shows the last 6 rows

* Names:
    * `names()` - returns the column names (synonym of `colnames()` for `data.frame`
       objects)
    * `rownames()` - returns the row names

* Summary:
    * `str()` - structure of the object and information about the class, length and
       content of  each column
    * `summary()` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.



## Maybe download the file first

It probably is best to first download the file, so that we have a
local copy, and then read it in.

We could first use `download_file()` to download the data into the `CleanData/` subdirectory:

```{r download_file, eval=FALSE, purl=TRUE}
download.file("http://kbroman.org/datacarp/portal_data_joined.csv",
              "CleanData/portal_data_joined.csv")
```

We could then use `read.csv()` to load the data into R, from the
downloaded file:

```{r read_csv_local_file, eval=FALSE}
surveys <- read.csv('CleanData/portal_data_joined.csv')
```



## Indexing, Sequences, and Subsetting

We can pull out parts of a data frame using square brackets.  We need
to provide two values: row and column, with a comma between them.

For example, to get the element in the 1st row, 1st column:

```{r one_one_element}
surveys[1,1]
```

To get the element in the 2nd row, 7th column:

```{r two_seven_element}
surveys[2,7]
```

To get the entire 2nd row, leave the column part blank:

```{r second_row}
surveys[2,]
```

And to get the entire 7th column, leave the row part blank:

```{r seventh_column}
sex <- surveys[,7]
```

You can also refer to columns by name, in multiple ways.

```{r grab_sex, eval = FALSE}
sex <- surveys[, "sex"]
sex <- surveys$sex
sex <- surveys[["sex"]]
```

When we pull out a single column, the result is a "vector". We can
again use square brackets to pull out individual values, but providing
a single number.

```{r grab_individual_sexes}
sex[1]
sex[10000]
```


### Slices

You can pull out larger slices from the vector by providing vectors of
indices.  You can use the `c()` function to create a vector.

```{r grab_some_sexes}
c(1,3,5)
sex[c(1,3,5)]
```

To pull out larger slices, it's
helpful to have ways of creating sequences of numbers.

First, the operator `:` gives you a sequence of consecutive values.

```{r}
1:10
10:1
5:8
```

`seq` is more flexible.

```{r}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
seq(10, 2, by=-2)  # can also go backwards
```


To get slices of our data frame, we can include a vector for the row
or column indexes (or both)

```{r}
surveys[1:3, 7]   # first three elements in the 7th column
surveys[1, 1:3]   # first three columns in the first row
surveys[2:4, 6:7] # rows 2-4, columns 6-7
```


### Challenge

The function `nrow()` on a `data.frame` returns the number of rows.

Use `nrow()`, in conjuction with `seq()` to create a new `data.frame` called
`surveys_by_10` that includes every 10th row of the survey data frame
starting at row 10 (10, 20, 30, ...)

<!-- end challenge -->





## Missing data

As R was designed to analyze datasets, it includes the concept of missing data
(which is uncommon in other programming languages). Missing data are represented
in vectors as `NA`.

```{r create_vec_with_missing}
heights <- c(2, 4, 4, NA, 6)
```

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values. It is a safer behavior as otherwise
you may overlook that you are dealing with missing data. You can add the
argument `na.rm=TRUE` to calculate the result while ignoring the missing values.

```{r na_rm}
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the
functions `is.na()` and `na.omit()`.


```{r extract_non_missing}
## Extract those elements which are not missing values.
heights[!is.na(heights)]
## shortcut to that
na.omit(heights)
```


### Treating blanks as missing

Another aside: it's probably best to treat those blanks as missing
(`NA`). To do that, use the argument `na.strings` when reading the
data. `na.strings` can be a vector of multiple character strings. We
need that a missing value code can never exist as a valid value,
because they all will be converted to the missing data code `NA`.  And
note that the default for `na.strings` is `"NA"`, which will cause
problems if `"NA"` is a valid value for your data (e.g., as an
abbreviation `"North America"`).

```{r read.csv_blanks_missing, eval=FALSE}
surveys_noblanks <- read.csv("CleanData/portal_data_joined.csv", na.strings="")
```



## Factors

Factors are used to represent categorical data. Factors can be ordered or
unordered, and understanding them is necessary for statistical analysis and for
plotting.

Factors are stored as integers, and have labels associated with these unique
integers. While factors look (and often behave) like character vectors, they are
actually integers under the hood, and you need to be careful when treating them
like strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. For
instance, if you use `factor()` to create a factor with 2 levels:

```{r create_sex_vector, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r levels_of_sex}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify a particular order.

```{r food_factor, purl=TRUE}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels=c("low", "medium", "high"))
levels(food)
```


### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

Converting factors where the levels appear as numbers (such as concentration
levels) to a numeric vector is a little trickier.  One method is to convert
factors to characters and then numbers.
function. Compare:

```{r convert_factor, purl=TRUE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
```

### Challenge

The function `table()` tabulates observations.

```{r table, eval=FALSE}
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1",
          "control", "treat1", "treat2", "treat3")
expt <- factor(expt)
table(expt)
```

* In which order are the treatments listed?
* How can you recreate this table with "`control`" listed last instead
of first?

<!-- end challenge -->

<!---
```{r correct-order}
## Answers
##
## * The treatments are listed in alphabetical order because they are factors.
## * By redefining the order of the levels
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1",
          "control", "treat1", "treat2", "treat3")
expt <- factor(expt, levels=c("treat1", "treat2", "treat3", "control"))
table(expt)
```
--->

### stringsAsFactors

The default when reading in data with `read.csv()`, columns with text
get turned into factors.

You can avoid this with the argument `stringsAsFactors=FALSE`.

```{r read_csv_stringsAsFactors, eval=FALSE}
surveys_chr <- read.csv("CleanData/portal_data_joined.csv", stringsAsFactors=FALSE)
```

Then when you look at the result of `str()`, you'll see that the
previously factor columns are now `chr`.

```{r str_on_that, eval=FALSE}
str(surveys_chr)
```
