---
layout: topic
title: Before we start (notes)
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

## Assigment operator

```{r assign}
age <- 2016 - 1969
```

- Paste into console.
- Can also use `=`, but recommend `<-`.
- Can use <kbd>Alt</kbd> + <kbd>-</kbd>

### Using the stuff you assigned

```{r sqrt}
sqrt(age)
```

```{r more_assign}
sqrt_age <- sqrt(age)
round(sqrt_age)
round(sqrt_age, 2)
```

### Object names

- explicit but not too long
- can't start with a number
- there are some names not allowed (`if`, `for`, `else`)
- other names you can but shouldn't use (`c`, `data`, `mean`)
- Can use `.`, but best not to

### Challenge

What is the value of `y` after doing the following?

```{r, purl=FALSE}
x <- 55
y <- x * 2.2
x <- 100
```

### Objects in workspace

- `ls()`
- Environment panel

## Vectors and data types

```{r vectors}
weights <- c(30, 100, 4000, 8000)
animals <- c("rat", "cat")
```

### length, class, str

```{r length}
length(weights)
length(animals)
```

```{r class_str}
class(weights)
class(animals)
str(weights)
str(animals)
```

### Add to a vector

```{r, purl=FALSE}
animals <- c(animals, "dog") # adding at the end of the vector
animals <- c("mouse", animals) # adding at the beginning of the vector
```


### Challenge

What happens to vectors with mixed types?
(Hint: use `class()`)

```{r, echo=TRUE, eval=FALSE, purl=TRUE}
x <- c(1, 2, 3, 'a')

y <- c(1, 2, 3, TRUE)

z <- c('a', 'b', 'c', TRUE)

tricky <- c(1, 2, 3, '4')
```

## Subsetting vectors

```{r, subset1}
animals <- c("mouse", "rat", "dog", "cat")
animals[2]
animals[c(1, 4)]
```

```{r, results='show', purl=FALSE}
more_animals <- animals[c(1, 2, 3, 2, 1, 4)]
```

### Challenge

Consider the `animals` vector.

```{r eval=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
```

Subset to get the 2nd and 3rd values.


### Conditional subsetting

Subsetting with a logical vector.

```{r, results='show', purl=FALSE}
weights <- c(21, 54, 39, 17, 55)
weights[c(FALSE, TRUE, FALSE, FALSE, TRUE)]
```

Creating a logical vector.

```{r, results='show', purl=FALSE}
weights > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weights[weights > 50]
```

Using `&` and `|`.

```{r, results='show', purl=FALSE}
weights[weights < 30 | weights > 50]
weights[weights > 30 & weights < 50]
```

Using `%in%` with character strings.

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]
```

### Challenge

Consider the following two vectors.

```{r, eval=FALSE}
animals <- c("mouse", "rat", "cat", "dog")
weights <- c(30, 180, 4000, 8000)
```

Subset `animals` with `weights < 200`.


## Missing data

```{r, purl=FALSE}
heights <- c(2, 4, 4, NA, 6)
```

```{r, purl=FALSE}
mean(heights)
max(heights)
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

```{r, purl=FALSE}
## Extract those elements which are not missing values.
heights[!is.na(heights)]
## shortcut to that
na.omit(heights)
```

### Challenge

Why does the following give an error?

```{r mean_na, eval=FALSE}
v <- c(2, 4, 4, "NA", 6)
mean(v, na.rm=TRUE)
```

Why does the error message say the data are not numeric?


<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
