---
layout: topic
title: Before we start (notes)
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

## Assigment operator

```{r assign}
age <- 2016 - 1969
```

- Paste into console.
- Can also use `=`, but recommend `<-`.
- Can use <kbd>Alt</kbd> + <kbd>-</kbd>

### Using the stuff you assigned

```{r sqrt}
sqrt(age)
```

```{r more_assign}
sqrt_age <- sqrt(age)
round(sqrt_age)
round(sqrt_age, 2)
```

### Object names

- explicit but not too long
- can't start with a number
- there are some names not allowed (`if`, `for`, `else`)
- other names you can but shouldn't use (`c`, `data`, `mean`)
- Can use `.`, but best not to

### Challenge

What is the value of `y` after doing the following?

```{r, purl=FALSE}
x <- 55
y <- x * 2.2
x <- 100
```

### Objects in workspace

- `ls()`
- Environment panel

## Vectors and data types

```{r vectors}
weights <- c(30, 100, 4000, 8000)
animals <- c("rat", "cat")
```

### length, class, str

```{r length}
length(weights)
length(animals)
```

```{r class_str}
class(weights)
class(animals)
str(weights)
str(animals)
```

### Add to a vector

```{r, purl=FALSE}
animals <- c(animals, "dog") # adding at the end of the vector
animals <- c("mouse", animals) # adding at the beginning of the vector
```


### Challenge

What happens to vectors with mixed types?
(Hint: use `class()`)

```{r, echo=TRUE, eval=FALSE, purl=TRUE}
x <- c(1, 2, 3, 'a')

y <- c(1, 2, 3, TRUE)

z <- c('a', 'b', 'c', TRUE)

tricky <- c(1, 2, 3, '4')
```




<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
