---
layout: topic
title: Data and data frames
author: Data Carpentry contributors
minutes: 50
---

```{r, echo=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * load external data (CSV files) in memory using the survey table
>  (`surveys.csv`) as an example
> * explore the structure and the content of a data frame in R
> * understand what factors are and how to manipulate them
> * understand the concept of a `data.frame`
> * use sequences
> * know how to access any element of a `data.frame`

------------

## Presentation of the Survey Data

We will continue to look at the species and weight of animals caught
in plots in a study area in Arizona over time. The dataset is stored
as a CSV file: each row holds information for a single animal, and the
columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

The data are available at <http://kbroman.org/datacarp/portal_data_joined.csv>.

We first use `download_file()` to download the data into the `data/` subdirectory:

```{r, eval=FALSE, purl=TRUE}
download.file("http://kbroman.org/datacarp/portal_data_joined.csv",
              "data/portal_data_joined.csv")
```

We then use `read.csv()` to load the data into R:

```{r, eval=TRUE}
surveys <- read.csv('data/portal_data_joined.csv')
```

We can actually use `read.csv` to grab the data directly from the web,
but it's probably best to download a copy first.

```{r eval=FALSE}
surveys <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```

The data are stored in what's called a "data frame". It's a big
rectangle, with rows being observations and columns being
variables. The different columns can be different types (numeric,
character, etc.), but they're all the same length.

Use `head()` to view the first few rows.

```{r head}
head(surveys)
```

Use `tail()` to view the first few rows.

```{r tail}
tail(surveys)
```

Use `str()` to look at the structure of the data.

```{r str}
str(surveys)
```

### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?

<!-- end challenge -->

As you can see, many of the columns consist of integers, however, the columns
`species` and `sex` are of a special class called a `factor`. Before we learn
more about the `data.frame` class, let's talk about factors. They are very
useful but not necessarily intuitive, and therefore require some attention.


## Factors

Factors are used to represent categorical data. Factors can be ordered or
unordered, and understanding them is necessary for statistical analysis and for
plotting.

Factors are stored as integers, and have labels associated with these unique
integers. While factors look (and often behave) like character vectors, they are
actually integers under the hood, and you need to be careful when treating them
like strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts *levels* in alphabetical order. For
instance, if you use `factor()` to create a factor with 2 levels:

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because
`f` comes before `m`, even though the first element in this vector is
`"male"`). You can check this by using the function `levels()`, and check the
number of levels using `nlevels()`:

```{r}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify a particular order.

```{r, purl=TRUE}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels=c("low", "medium", "high"))
levels(food)
```


### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

Converting factors where the levels appear as numbers (such as concentration
levels) to a numeric vector is a little trickier.  One method is to convert
factors to characters and then numbers.
function. Compare:

```{r, purl=TRUE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
```

### Challenge

The function `table()` tabulates observations.

```{r table, eval=FALSE}
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1",
          "control", "treat1", "treat2", "treat3")
expt <- factor(expt)
table(expt)
```

* In which order are the treatments listed?
* How can you recreate this table with "`control`" listed last instead
of first?

<!-- end challenge -->

<!---
```{r correct-order}
## Answers
##
## * The treatments are listed in alphabetical order because they are factors.
## * By redefining the order of the levels
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1",
          "control", "treat1", "treat2", "treat3")
expt <- factor(expt, levels=c("treat1", "treat2", "treat3", "control"))
table(expt)
```
--->

## stringsAsFactors

The default when reading in data with `read.csv()`, columns with text
get turned into factors.

You can avoid this with the argument `stringsAsFactors=FALSE`.

```{r}
surveys_chr <- read.csv("data/portal_data_joined.csv", stringsAsFactors=FALSE)
```

Then when you look at the result of `str()`, you'll see that the
previously factor columns are now `chr`.

```{r}
str(surveys_chr)
```

## Constructing data frames "by hand"

You can also create a data frame manually with the function `data.frame()`. This
function can also take the argument `stringsAsFactors`. Compare the output of
these examples, and compare the difference between when the data are being read
as `character`, and when they are being read as `factor`.

```{r, results='show'}
df1 <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                  feel=c("furry", "furry", "squishy", "spiny"),
                  weight=c(45, 8, 1.1, 0.8))
str(df1)

df2 <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                  feel=c("furry", "furry", "squishy", "spiny"),
                  weight=c(45, 8, 1.1, 0.8), stringsAsFactors=FALSE)
str(df2)
```

### Challenge

There are a few mistakes in this hand crafted `data.frame`, can you spot and
fix them? Don't hesitate to experiment!

```{r, eval=FALSE}
author_book <- data.frame(author_first=c("Charles", "Ernst", "Theodosius"),
                          author_last=c(Darwin, Mayr, Dobzhansky),
                          year=c(1942, 1970))
```

<!-- end challenge -->

## Inspecting data frames

We already saw how the functions `head()` and `str()` can be useful to check the
content and the structure of a `data.frame`. Here is a non-exhaustive list of
functions to get a sense of the content/structure of the data.

* Size:
    * `dim()` - returns a vector with the number of rows in the first element,
      and the number of columns as the second element (the __dim__ensions of the
      object)
    * `nrow()` - returns the number of rows
    * `ncol()` - returns the number of columns

* Content:
    * `head()` - shows the first 6 rows
    * `tail()` - shows the last 6 rows

* Names:
    * `names()` - returns the column names (synonym of `colnames()` for `data.frame`
       objects)
    * `rownames()` - returns the row names

* Summary:
    * `str()` - structure of the object and information about the class, length and
       content of  each column
    * `summary()` - summary statistics for each column

Note: most of these functions are "generic", they can be used on other types of
objects besides `data.frame`.

## Indexing, Sequences, and Subsetting

We pulled out parts of a vector by indexing with square brackets. We
can do the same thing with data frames, but we need to provide two
values: row and column, with a comma between them.

For example, to get the element in the 1st row, 1st column:

```{r}
surveys[1,1]
```

To get the element in the 2nd row, 7th column:

```{r}
surveys[2,7]
```

To get the entire 2nd row, leave the column part blank:

```{r}
surveys[2,]
```

And to get the entire 7th column, leave the row part blank:

```{r}
sex <- surveys[,7]
```

You can also refer to columns by name, in multiple ways.

```{r, eval = FALSE}
sex <- surveys[, "sex"]
ssex <- surveys[["sex"]]
sex <- surveys$sex
```

### Treating blanks as missing

Another aside: it's probably best to treat those blanks as missing
(`NA`). To do that, use the argument `na.strings` when reading the
data.

```{r}
surveys_noblanks <- read.csv("data/portal_data_joined.csv", na.strings=c("", "NA"))
```


### Slices

As with vectors, you can also use logical vectors when indexing.

```{r}
weights_males <- surveys[surveys$sex == 'M', "weight"]
mean(weights_males, na.rm=TRUE)

mean(surveys[surveys$sex == 'F', "weight"], na.rm=TRUE)
```

Or you can use numeric vectors. To pull out larger slices, it's
helpful to have ways of creating sequences of numbers.

First, the operator `:` gives you a sequence of consecutive values.


```{r}
1:10
10:1
5:8
```

`seq` is more flexible.

```{r}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
seq(10, 2, by=-2)  # can also go backwards
```




To get slices of our data frame, we can include a vector for the row
or column indexes (or both)

```{r}
surveys[1:3, 7]   # first three elements in the 7th column
surveys[1, 1:3]   # first three columns in the first row
surveys[2:4, 6:7] # rows 2-4, columns 6-7
```


### Challenge

1. The function `nrow()` on a `data.frame` returns the number of rows. Use it,
   in conjuction with `seq()` to create a new `data.frame` called
   `surveys_by_10` that includes every 10th row of the survey data frame
   starting at row 10 (10, 20, 30, ...)

1. Create a `data.frame` containing only the observations from the
   year 1999 of the `surveys` dataset.

<!-- end challenge -->
