---
layout: topic
title: Starting with data (notes)
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

## Load the data

We first download the file:

```{r, eval=FALSE, purl=TRUE}
download.file("http://kbroman.org/datacarp/portal_data_joined.csv",
              "data/portal_data_joined.csv")
```

We then load them into R:

```{r, eval=TRUE, purl=FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
```

Can also do:

```{r eval=FALSE}
surveys <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```

Use `head()` to view the first few rows.

```{r head}
head(surveys)
```

Use `tail()` to view the last few rows.


```{r tail}
tail(surveys)
```

Use `str()` to look at the structure of the data.

```{r str}
str(surveys)
```

### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?


## Factors

Factors are categorical variables. They look like a vector of
character strings, but they're stored as integers with character
string labels for the distinct values, called _levels_.

Use `factor()` to create a vector that is a factor.

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

Use `levels()` and `nlevels()` to see the levels and their number.

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

If you provide the `levels` argument to `factor()`, you'll get the
levels in a particular order.

```{r, purl=TRUE}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels=c("low", "medium", "high"))
levels(food)
```

Converting factors to numeric.

```{r, purl=TRUE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
```

### Challenge

The function `table()` tabulates observations.

```
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1", "control",
          "treat1", "treat2", "treat3")
expt <- factor(expt)
table(expt)
```

* In which order are the treatments listed?
* How can you recreate this table with "control" listed last instead
of first?


<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
