---
layout: topic
title: Starting with data (notes)
author: Data Carpentry contributors
minutes: 15
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

## Load the data

We first download the file:

```{r, eval=FALSE, purl=TRUE}
download.file("http://kbroman.org/datacarp/portal_data_joined.csv",
              "data/portal_data_joined.csv")
```

We then load them into R:

```{r, eval=TRUE, purl=FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
```

Can also do:

```{r eval=FALSE}
surveys <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv")
```

Use `head()` to view the first few rows.

```{r head}
head(surveys)
```

Use `tail()` to view the last few rows.


```{r tail}
tail(surveys)
```

Use `str()` to look at the structure of the data.

```{r str}
str(surveys)
```

### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?


## Factors

Factors are categorical variables. They look like a vector of
character strings, but they're stored as integers with character
string labels for the distinct values, called _levels_.

Use `factor()` to create a vector that is a factor.

```{r, purl=TRUE}
sex <- factor(c("male", "female", "female", "male"))
```

Use `levels()` and `nlevels()` to see the levels and their number.

```{r, purl=FALSE}
levels(sex)
nlevels(sex)
```

If you provide the `levels` argument to `factor()`, you'll get the
levels in a particular order.

```{r, purl=TRUE}
food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
levels(food)
food <- factor(food, levels=c("low", "medium", "high"))
levels(food)
```

Converting factors to numeric.

```{r, purl=TRUE}
f <- factor(c(1, 5, 10, 2))
as.numeric(f)               ## wrong! and there is no warning...
as.numeric(as.character(f)) ## works...
```

### Challenge

The function `table()` tabulates observations.

```
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1", "control",
          "treat1", "treat2", "treat3")
expt <- factor(expt)
table(expt)
```

* In which order are the treatments listed?
* How can you recreate this table with "control" listed last instead
of first?


<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
---
layout: topic
title: Data frames (notes)
author: Data Carpentry contributors
minutes: 30
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
surveys <- read.csv("data/portal_data_joined.csv")
```

## stringsAsFactors

Avoid factors with `stringsAsFactors=FALSE`.

```{r}
surveys_chr <- read.csv("data/portal_data_joined.csv", stringsAsFactors=FALSE)
```

```{r}
str(surveys_chr)
```


## Constructing data frames "by hand"

```{r, results='show'}
df1 <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                  feel=c("furry", "furry", "squishy", "spiny"),
                  weight=c(45, 8, 1.1, 0.8))
str(df1)

df2 <- data.frame(animal=c("dog", "cat", "sea cucumber", "sea urchin"),
                  feel=c("furry", "furry", "squishy", "spiny"),
                  weight=c(45, 8, 1.1, 0.8), stringsAsFactors=FALSE)
str(df2)
```


### Challenge

There are a few mistakes in this hand crafted `data.frame`, can you spot and
fix them? Don't hesitate to experiment!

```{r, eval=FALSE}
author_book <- data.frame(author_first=c("Charles", "Ernst", "Theodosius"),
                          author_last=c(Darwin, Mayr, Dobzhansky),
                          year=c(1942, 1970))
```

<!-- end challenge -->


## Inspecting data frames

```{r}
dim(surveys)
nrow(surveys)
ncol(surveys)

names(surveys)
rownames(surveys)

summary(surveys)
```

## Indexing, sequences, subsetting

```{r}
surveys[1,1]

surveys[2,7]

surveys[2,]

sex <- surveys[,7]
```

Referring to columns.

```{r, eval = FALSE}
species_id <- surveys[, "sex"]
species_id <- surveys[["sex"]]
species_id <- surveys$sex
```


### Treating blanks as missing

```{r}
surveys_noblanks <- read.csv("data/portal_data_joined.csv", na.strings=c("", "NA"))
```

### Slices

As with vectors, you can also use logical vectors when indexing.

```{r}
weights_males <- surveys[surveys$sex == 'M', "weight"]
mean(weights_males, na.rm=TRUE)

mean(surveys[surveys$sex == 'F', "weight"], na.rm=TRUE)
```

Or you can use numeric vectors. To pull out larger slices, it's
helpful to have ways of creating sequences of numbers.

First, `:`


```{r}
1:10
10:1
5:8
```

`seq` is more flexible.

```{r}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
seq(10, 2, by=-2)  # can also go backwards
```

Use these to take slices of the data.

```{r}
surveys[1:3, 7]   # first three elements in the 7th column
surveys[1, 1:3]   # first three columns in the first row
surveys[2:4, 6:7] # rows 2-4, columns 6-7
```

### Challenge

1. The function `nrow()` on a `data.frame` returns the number of rows. Use it,
   in conjuction with `seq()` to create a new `data.frame` called
   `surveys_by_10` that includes every 10th row of the survey data frame
   starting at row 10 (10, 20, 30, ...)

2. Create a `data.frame` containing only the observations from 1999 of the
   `surveys` dataset.



<br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/> <br/>
