---
output: slidy_presentation
---

# Challenge 1


What is the value of `y` after doing the following?

```{r, eval=FALSE}
x <- 55
y <- x * 2.2
x <- 100
```


---

# Challenge 2


What happens to vectors with mixed types?

```{r, echo=TRUE, eval=FALSE}
x <- c(1, 2, 3, 'a')

y <- c(1, 2, 3, TRUE)

z <- c('a', TRUE, 'b', 'c')

tricky <- c(1, '2', 3, 4)
```

Hint: use `class()`


---

# Challenge 3


Consider the `animals` vector.

```{r eval=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
```

Subset to get the 2nd and 3rd values.


### Conditional subsetting

Another common way of subsetting is by using a logical vector: `TRUE` will
select the element with the same index, while `FALSE` will not:

```{r, results='show', purl=FALSE}
weights <- c(21, 54, 39, 17, 55)
weights[c(FALSE, TRUE, FALSE, FALSE, TRUE)]
```

Typically, these logical vectors are not typed by hand, but are the output of
other functions or logical tests. For instance, if you wanted to select only the
values above 50:

```{r, results='show', purl=FALSE}
weights > 50    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 50
weights[weights > 50]
```

You can combine multiple tests using `&` (both conditions are true, AND) or `|`
(at least one of the conditions is true, OR):

```{r, results='show', purl=FALSE}
weights[weights < 30 | weights > 50]
weights[weights > 30 & weights < 50]
```

When working with vectors of characters, if you are trying to combine many
conditions it can become tedious to type. The function `%in%` allows you to test
if a value if found in a vector:

```{r, results='show', purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]
```

### Challenge

Consider the following two vectors.

```{r, eval=FALSE}
animals <- c("mouse", "rat", "cat", "dog")
weights <- c(30, 180, 4000, 8000)
```

Subset `animals` with `weights < 200`.


---

# Challenge 4


Why does the following give an error?

```{r mean_na, eval=FALSE}
v <- c(2, 4, 4, "NA", 6)
mean(v, na.rm=TRUE)
```

Why does the error message say the data are not numeric?


