---
output: slidy_presentation
---

# Challenge 1


What is the value of `y` after doing the following?

```{r, eval=FALSE}
x <- 55
y <- x * 2.2
x <- 100
```


---

# Challenge 2


What happens to vectors with mixed types?

```{r, echo=TRUE, eval=FALSE}
x <- c(1, 2, 3, 'a')

y <- c(1, 2, 3, TRUE)

z <- c('a', TRUE, 'b', 'c')

tricky <- c(1, '2', 3, 4)
```

Hint: use `class()`


---

# Challenge 3


Consider the `animals` vector.

```{r eval=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
```

Subset to get the 2nd and 3rd values.


---

# Challenge 4


Consider the following two vectors.

```{r, eval=FALSE}
animals <- c("mouse", "rat", "cat", "dog")
weights <- c(30, 180, 4000, 8000)
```

Subset `animals` with `weights < 200`.


---

# Challenge 5


Why does the following give an error?

```{r mean_na, eval=FALSE}
v <- c(2, 4, 4, "NA", 6)
mean(v, na.rm=TRUE)
```

Why does the error message say the data are not numeric?


---

# Challenge 6


Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?


---

# Challenge 7


The function `table()` tabulates observations.

```{r table, eval=FALSE}
expt <- c("treat1", "treat2", "treat1", "treat3", "treat1",
          "control", "treat1", "treat2", "treat3")
expt <- factor(expt)
table(expt)
```

* In which order are the treatments listed?
* How can you recreate this table with "`control`" listed last instead
of first?


---

# Challenge 8


There are a few mistakes in this hand crafted `data.frame`, can you spot and
fix them? Don't hesitate to experiment!

```{r, eval=FALSE}
author_book <- data.frame(author_first=c("Charles", "Ernst", "Theodosius"),
                          author_last=c(Darwin, Mayr, Dobzhansky),
                          year=c(1942, 1970))
```


---

# Challenge 9


1. The function `nrow()` on a `data.frame` returns the number of rows. Use it,
   in conjuction with `seq()` to create a new `data.frame` called
   `surveys_by_10` that includes every 10th row of the survey data frame
   starting at row 10 (10, 20, 30, ...)

1. Create a `data.frame` containing only the observations from the
   year 1999 of the `surveys` dataset.


---

# Challenge 10


Using pipes, subset the data to include individuals collected before 1995,
and retain the columns `year`, `sex`, and `weight.`


---

# Challenge 11


Create a new dataframe from the survey data that meets the following
criteria: contains only the `species_id` column and a column that contains
values that are half the `hindfoot_length` values (e.g. a new column
`hindfoot_half`). In this `hindfoot_half` column, there are no NA values
and all values are < 30.

Hint: think about how the commands should be ordered to produce this data frame!


---

# Challenge 12


How many times was each `plot_type` surveyed?


---

# Challenge 13


Use `group_by()` and `summarize()` to find the mean, min, and max hindfoot
length for each species.


---

# Challenge 14


What was the heaviest animal measured in each year? Return the columns `year`,
`genus`, `species`, and `weight`.


---

# Challenge 15


Make a scatterplot of `hindfoot_length` vs `weight`, but only for the
`species_id`, `"DM"`.


---

# Challenge 16


Use dplyr to calculate the mean `weight` and `hindfoot_length` as well
as the sample size for each species.

Make a scatterplot of mean `hindfoot_length` vs `mean_weight`, with
the sizes of the points corresponding to the sample size.


---

# Challenge 17


Make a plot of counts of `species_id` `"DM"` and `"DS"` by year. See
if you can make the lines black but the points colored by species.


---

# Challenge 18


Try using `geom_histogram()` to make a histogram visualization of the
distribution of `weight`.

Hint: You want `weight` as the x-axis aesthetic. Try specifying `bins`
in `geom_histogram()`.


---

# Challenge 19


A variant on the box plot is the violin plot. Use `geom_violin()` to
make violin plots of `hindfoot_length` by `species_id`.


---

# Challenge 20


- Calculate counts grouped by year, species_id, and sex

- make the faceted plot splitting further by sex (within each panel)

- color by sex rather than species


---

# Challenge 21


- Create a new R Markdown document.
- Delete all of the R code chunks and write a bit of Markdown (some sections, some italicized
text, and an itemized list).
- Convert the document to a webpage.


---

# Challenge 22


Add code chunks to

- Load the ggplot2 package
- Read the portal data
- Create a plot


---

# Challenge 23


Use chunk options to control the size of a figure and to hide the
code.


---

# Challenge 24


Try out a bit of in-line R code.



# Capstone project

Create and compile an R Markdown report:

1. Load the `combined.csv` data.

2. Create boxplots of weight by sex, _omitting individuals with
missing sex_.

3. Create a histogram of hindfoot lengths.

4. Create a scatterplot of hindfoot length vs weight for the species
`"DM"`, `"DO"`, and `"DS"`. _Use different colors for the three
species, and put the three species in different panels._

5. Create a table of counts of `"DM"` by plot type for the year 1977.

6. Create a line plot of the counts of `"DM"` in `"Rodent Exclosure"` plots over time.
