---
output: slidy_presentation
---

# Challenge 1


What is the value of `y` after doing the following?

```{r, eval=FALSE}
x <- 55
y <- x * 2.2
x <- 100
```


---

# Challenge 2


What happens to vectors with mixed types?

```{r, echo=TRUE, eval=FALSE}
x <- c(1, 2, 3, 'a')

y <- c(1, 2, 3, TRUE)

z <- c('a', TRUE, 'b', 'c')

tricky <- c(1, '2', 3, 4)
```

Hint: use `class()`


---

# Challenge 3


Consider the `animals` vector.

```{r eval=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
```

Subset to get the 2nd and 3rd values.


---

# Challenge 4


Consider the following two vectors.

```{r, eval=FALSE}
animals <- c("mouse", "rat", "cat", "dog")
weights <- c(30, 180, 4000, 8000)
```

Subset `animals` with `weights < 200`.


---

# Challenge 5


Why does the following give an error?

```{r mean_na, eval=FALSE}
v <- c(2, 4, 4, "NA", 6)
mean(v, na.rm=TRUE)
```

Why does the error message say the data are not numeric?



# Capstone project

Create and compile an R Markdown report:

1. Load the `combined.csv` data.

2. Create boxplots of weight by sex, _omitting individuals with
missing sex_.

3. Create a histogram of hindfoot lengths.

4. Create a scatterplot of hindfoot length vs weight for the species
`"DM"`, `"DO"`, and `"DS"`. _Use different colors for the three
species, and put the three species in different panels._

5. Create a table of counts of `"DM"` by plot type for the year 1977.

6. Create a line plot of the counts of `"DM"` in `"Rodent Exclosure"` plots over time.
